---
title: "R Notebook"
output: html_notebook
---

# Final Project for INFO 3010

## AI/ML Salary Predictor

## Maya Espinosa

### Open and Check the Dataset

```{r}
#import the data
salaries <- read.csv("C:/Users/Mayae/OneDrive/Documents/UNT/Spring 2024/DTSC 3010 Intro to D.S/salaries.csv")
```

```{r}
#check the dataset
head(salaries)
```

```{r}
summary(salaries)
```

```{r}
str(salaries)
```

### Libraries for Data Cleaning and EDA's

```{r}
# Load necessary libraries for data cleaning and EDA's
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
```

#### Data Cleaning

```{r}
# Remove rows with null or zero salary values
salaries <- subset(salaries, !is.na(salary_in_usd) & salary_in_usd != 0)
# Check the dimensions of the dataset after removing rows
cat("Dimensions of the dataset after removing rows:", nrow(salaries), "rows and", ncol(salaries), "columns.")
#there was no missing values
```

```{r}
# Check if any job titles include "internship" or "intern"
internship_jobs <- subset(salaries, grepl("internship|intern", job_title, ignore.case = TRUE))
# Display the job titles containing "internship" or "intern"
if (nrow(internship_jobs) > 0) {
  cat("Job titles containing 'internship' or 'intern':\n")
  cat(internship_jobs$job_title, "\n")
} else {
  cat("No job titles contain 'internship' or 'intern'.\n")
}
```

```{r}
summary(salaries$salary_in_usd)
```

```{r}
#Which jobs make less than $25,000 USD (low range of data engineer)
# Filter rows with salary less than $25,000 USD
low_salary_jobs <- subset(salaries, salary_in_usd < 25000)
# Drop the columns "salary" and "salary_currency"
low_salary_jobs <- low_salary_jobs[, !(names(low_salary_jobs) %in% c("salary", "salary_currency"))]
# Display the job titles of the jobs with low salaries
if (nrow(low_salary_jobs) > 0) {
  cat("Job titles of jobs with salary less than $25,000 USD:")
  cat(unique(low_salary_jobs$job_title))
} else {
  cat("No jobs with salary less than $25,000 USD.")
}
# Display the table of jobs with low salaries
if (nrow(low_salary_jobs) > 0) {
  print(low_salary_jobs)
} else {
  cat("No jobs with salary less than $25,000 USD.")
}
```

```{r}
# Create a histogram of salaries from low_salary_jobs
ggplot(low_salary_jobs, aes(x = salary_in_usd, fill = ..count..)) +
  geom_histogram(color = "black", bins = 20) +
  theme_minimal() +
  labs(title = "Histogram of Salaries (<$20,000 USD)",
       x = "Salary (USD)",
       y = "Frequency") +
  scale_fill_gradient("Frequency", low = "#fbc046", high = "red")
```

#### Exploratory Data Analysis (EDA)

```{r}
# Histogram of salaries
ggplot(salaries, aes(x = salary_in_usd)) +
  geom_histogram(binwidth = 5000, fill = "#cb6ce6", color = "black") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(title = "Distribution of Salaries",
       x = "Salary (USD)",
       y = "Frequency") +
  theme_minimal()
```

```{r}
summary(salaries)
```

```{r}
# Box plot of salaries by job title
ggplot(salaries, aes(x = job_title, y = salary_in_usd)) +
  geom_boxplot(fill = "#0f52fb", color = "black") +
  labs(title = "Salary Distribution by Job Title",
       x = "Job Title",
       y = "Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)
```

```{r}
# Bar plot of job title frequencies
job_title_counts <- salaries %>% 
  count(job_title) %>%
  arrange(desc(n))

ggplot(job_title_counts, aes(x = reorder(job_title, -n), y = n)) +
  geom_bar(stat = "identity", fill = "#f9b61a", color = "black") +
  labs(title = "Frequency of Job Titles",
       x = "Job Title",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#fix to make it more readable (change the filter value as needed)
# Filter out job titles with n frequency
filtered_job_title_counts <- job_title_counts %>%
  filter(n > 100)

ggplot(filtered_job_title_counts, aes(x = reorder(job_title, -n), y = n)) +
  geom_bar(stat = "identity", fill = "#f9b61a", color = "black") +
  scale_y_continuous(labels = scales::comma) +  # Format y-axis labels with commas
  labs(title = "Frequency of Job Titles",
       x = "Job Title",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Correlation matrix
correlation_matrix <- cor(salaries[, c("salary_in_usd", "remote_ratio")])
# Convert correlation matrix to long format
correlation_long <- as.data.frame(as.table(correlation_matrix))
names(correlation_long) <- c("Var1", "Var2", "value")
# Heatmap of correlation matrix
ggplot(data = correlation_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "#f9b61a", high = "#f52e18") +
  labs(title = "Correlation Heatmap",
       x = "",
       y = "")

```

```{r}
# Scatter plot of salary vs. years of experience
ggplot(salaries, aes(x = experience_level, y = salary_in_usd)) +
  geom_point(alpha = 0.5, color = "#cb6ce6") +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-3)) +  # Format y-axis labels in thousands
  labs(title = "Salary vs. Years of Experience",
       x = "Experience Level",
       y = "Salary (USD in Thousands)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Calculate frequency of each unique remote_ratio value
remote_ratio_counts <- table(salaries$remote_ratio)
# Convert the counts to a data frame for plotting
remote_ratio_df <- data.frame(remote_ratio = as.numeric(names(remote_ratio_counts)),
                              count = as.numeric(remote_ratio_counts))
# Define pink color palette
pink_palette <- c("#cb6ce6", "#0f52fb", "#b2de55")
# Plot pie chart
ggplot(remote_ratio_df, aes(x = "", y = count, fill = factor(remote_ratio))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Distribution of Remote Work Ratio",
       fill = "Remote Work",
       x = NULL,
       y = NULL) +
  scale_fill_manual(values = pink_palette) +  #
  theme_void() +
  theme(legend.position = "right")
```

### Statistical Tests

```{r}
#Load required packages
library(stats)
```

```{r}
# Run ANOVA test
anova_result <- aov(salary_in_usd ~ experience_level, data = salaries)

# Summary of ANOVA results
print("One-Way ANOVA:")
summary(anova_result)
```

-   The **`experience_level`** variable has a statistically significant effect on **`salary_in_usd`** (p \< 0.001).

-   This suggests that there are significant differences in **`salary_in_usd`** across different levels of **`experience_level`**.

```{r}
# Perform Pearson's chi-square test
chi_square_result <- chisq.test(table(salaries$employment_type, salaries$remote_ratio))

# Summary of chi-square results
print(chi_square_result)
```

-   **X-squared**: The chi-squared statistic value is 140.28.

-   **Degrees of Freedom (df)**: The degrees of freedom associated with the chi-squared statistic is 6.

-   **p-value**: The p-value is less than 2.2e-16 (essentially zero), suggesting strong evidence against the null hypothesis. Thus, we reject the null hypothesis of independence.

```{r}
# Pearson Correlation Test
correlation_result <- cor.test(salaries$salary_in_usd, salaries$remote_ratio, method = "pearson")
print("Pearson Correlation Test:")
print(correlation_result)
```

-   The correlation coefficient between **`salary_in_usd`** and **`remote_ratio`** is approximately -0.095. This suggests a weak negative correlation between the two variables.

-   The p-value is very small (2.62e-16), indicating strong evidence against the null hypothesis of no correlation.

-   Therefore, we reject the null hypothesis and conclude that there is a significant correlation between **`salary_in_usd`** and **`remote_ratio`**.

```{r}
# Perform ANOVA test
anova_result <- aov(salary_in_usd ~ work_year, data = salaries)

# Summary of ANOVA test
summary(anova_result)
```

-   The F value is 310.2, indicating a significant association between **`work_year`** and **`salary_in_usd`**.

-   The p-value (**`Pr(>F)`**) is less than 2e-16, indicating that the association between **`work_year`** and **`salary_in_usd`** is statistically significant.

```{r}
# Perform ANOVA
anova_result <- aov(salary_in_usd ~ job_title, data = salaries)
print(summary(anova_result))

# Perform Kruskal-Wallis test
kruskal_test_result <- kruskal.test(salary_in_usd ~ job_title, data = salaries)
print(kruskal_test_result)
```

-   In the ANOVA test, the p-value is less than 0.05, indicating that there are significant differences in the mean salaries across different job titles.

-   Similarly, in the Kruskal-Wallis test, the p-value is less than 0.05, suggesting that there are significant differences in the median salaries across different job titles.

```{r}
anova_result <- aov(salary_in_usd ~ work_year + employment_type + experience_level + job_title, data = salaries)
print(summary(anova_result))
```

-   Overall, this ANOVA suggests that normalized work year, experience level, job title, and employment type are all significantly associated with variations in salary within your dataset.

    ### Supervised Learning

```{r}
# Fit linear regression model
linear_model <- lm(salary_in_usd ~ work_year + employment_type + experience_level + job_title, data = salaries)

# Summary of the linear regression model
summary(linear_model)

```

-   **Adjusted R-squared**: The adjusted R-squared value is 0.3058, which indicates that approximately 30.58% of the variance in **`salary_in_usd`** can be explained by the combination of **`work_year`**, **`employment_type`**, **`experience_level`**, and **`job_title`**.

-   **Coefficients**:

    -   For **`work_year`**, **`employment_type`**, **`experience_level`**, and **`job_title`**, the coefficients represent the change in salary relative to the reference categories or levels. For example, compared to employees with full-time (FT) employment, employees with freelance (FL) employment have a salary approximately \$69800 lower, on average, holding other variables constant.

-   **Model Significance**: The overall model is statistically significant, as indicated by the F-statistic (F = 27.16, p \< 2.2e-16).

-   **Residuals**: The residual standard error is 53250, indicating the average deviation of the observed values from the fitted values.

```{r}
# Load the randomForest library
library(randomForest)

# Fit random forest model
rf_model <- randomForest(salary_in_usd ~ work_year + employment_type + experience_level + job_title, data = salaries)

# Summary of the random forest model
print(summary(rf_model))

# Get predicted values from the model
predicted_values <- predict(rf_model)

# Calculate Mean Squared Error (MSE)
mse <- mean((salaries$salary_in_usd - predicted_values)^2)

# Calculate R-squared (rsq)
actual_mean <- mean(salaries$salary_in_usd)
rsq <- 1 - sum((salaries$salary_in_usd - predicted_values)^2) / sum((salaries$salary_in_usd - actual_mean)^2)
print("Mean Squared Error:")
print(mse)
print("R-squared:")
print(rsq)
```

```{r}
# Load the e1071 library for SVM
library(e1071)

# Fit SVM regression model
svm_model <- svm(salary_in_usd ~ work_year + employment_type + experience_level + job_title, data = salaries)

# Summary of the SVM regression model
print(summary(svm_model))

# Get predicted values from the model
predicted_values <- predict(svm_model)

# Calculate R-squared (rsq)
actual_mean <- mean(salaries$salary_in_usd)
rsq <- 1 - sum((salaries$salary_in_usd - predicted_values)^2) / sum((salaries$salary_in_usd - actual_mean)^2)
print("R-squared:")
print(rsq)
```

```{r}
# Load the rpart library for regression trees
library(rpart)

# Fit regression tree model with specified cp value
tree_model <- rpart(salary_in_usd ~ work_year + employment_type + experience_level + job_title,
                    data = salaries,
                    cp = 0.001)  # Specify the desired cp value

# Summary of the regression tree model
print(summary(tree_model))

# Get predicted values from the model
predicted_values <- predict(tree_model)

# Calculate R-squared (rsq)
actual_mean <- mean(salaries$salary_in_usd)
rsq <- 1 - sum((salaries$salary_in_usd - predicted_values)^2) / sum((salaries$salary_in_usd - actual_mean)^2)
print(rsq)

```

```{r}
# Create a bar chart to show variable importance
variable_importance <- c(49, 43, 7, 1)  # Variable importance values
variables <- c("job_title", "experience_level", "work_year", "employment_type")  # Variable names

barplot(variable_importance, names.arg = variables, 
        col = "red", main = "Variable Importance in Regression Tree Model",
        xlab = "Variables", ylab = "Importance")

# Print the Mean Squared Error (MSE)
MSE <- 4.084373e+09  # Mean Squared Error value from the model summary
cat("Mean Squared Error (MSE):", MSE)

```

### Prediction

```{r}
# Load the rpart library for regression trees
library(rpart)

# Fit regression tree model with specified cp value
tree_model <- rpart(salary_in_usd ~ work_year + employment_type + experience_level + job_title,
                    data = salaries,
                    cp = 0.001)  # Specify the desired cp value

# Function to predict salary based on user inputs
predict_salary <- function() {
  # Get user inputs
  work_year <- as.numeric(readline("Enter work year: "))  # Input for work_year
  employment_type <- readline("Enter employment type (FT, PT, CT, FL): ")  # Input for employment_type
  experience_level <- readline("Enter experience level (EN, MI, SE, EX): ")  # Input for experience_level
  job_title <- readline("Enter job title: ")  # Input for job_title
  
  # Create a data frame with user inputs
  user_inputs <- data.frame(
    work_year = work_year,
    employment_type = employment_type,
    experience_level = experience_level,
    job_title = job_title
  )
  
  # Predict salary based on user inputs
  predicted_salary <- predict(tree_model, newdata = user_inputs)
  
  # Return predicted salary
  return(predicted_salary)
}

# Predict salary based on user inputs
predicted_salary <- predict_salary()

# Print predicted salary
cat("Predicted salary in USD: $", predicted_salary, "\n")

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
